---
title: "Advanced Features in CostUtilization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features in CostUtilization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Introduction ðŸš€

Ready to unlock the full power of the CostUtilization package? This vignette covers advanced features that will help you conduct sophisticated healthcare cost analyses. We'll explore complex filtering, multi-cohort comparisons, performance optimization, and more.

```{r setup}
library(CostUtilization)
library(DatabaseConnector)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(cli)

# Setup connection to Eunomia for examples
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
connection <- DatabaseConnector::connect(connectionDetails)

# Ensure cost data is available
injectCostData(connection, cdmDatabaseSchema = "main", seed = 123)
transformCostToCdmV5dot4(connectionDetails, cdmDatabaseSchema = "main")

# Set theme for plots
theme_set(theme_minimal() + 
          theme(plot.title = element_text(face = "bold"),
                plot.subtitle = element_text(color = "gray50")))
```

## Complex Event Filtering ðŸ”

### Building Comprehensive Event Filter Sets

Event filters let you analyze costs for specific clinical scenarios. Here's how to build sophisticated filter sets:

```{r complex-filters}
# Example: Comprehensive cardiovascular care pathway
cvCarePathway <- list(
  # === ACUTE EVENTS ===
  list(
    name = "Acute MI",
    domain = "Condition",
    conceptIds = c(4329847, 314666, 4108217, 4296653, 4185932)
  ),
  list(
    name = "Unstable Angina",
    domain = "Condition",
    conceptIds = c(321318, 4219032)
  ),
  list(
    name = "Heart Failure",
    domain = "Condition",
    conceptIds = c(316139, 319835, 321319)
  ),
  
  # === DIAGNOSTIC PROCEDURES ===
  list(
    name = "Cardiac Catheterization",
    domain = "Procedure",
    conceptIds = c(4019824, 4297650)
  ),
  list(
    name = "Echocardiography",
    domain = "Procedure",
    conceptIds = c(4305178, 4013419)
  ),
  list(
    name = "Stress Testing",
    domain = "Procedure",
    conceptIds = c(4143316, 4327941)
  ),
  
  # === INTERVENTIONS ===
  list(
    name = "PCI",
    domain = "Procedure",
    conceptIds = c(4336464, 4178904, 4124836)
  ),
  list(
    name = "CABG",
    domain = "Procedure",
    conceptIds = c(4336464, 4000733, 4228152)
  ),
  
  # === MEDICATIONS ===
  list(
    name = "Antiplatelet Agents",
    domain = "Drug",
    conceptIds = c(1308216, 1310149, 1322184, 1318011)  # Aspirin, clopidogrel, ticagrelor, prasugrel
  ),
  list(
    name = "Beta Blockers",
    domain = "Drug",
    conceptIds = c(1307046, 1313200, 1308842, 1345858)  # Metoprolol, atenolol, carvedilol, bisoprolol
  ),
  list(
    name = "ACE/ARB",
    domain = "Drug",
    conceptIds = c(1308216, 1334456, 1341927, 1367500)  # Lisinopril, losartan, valsartan, ramipril
  ),
  list(
    name = "Statins",
    domain = "Drug",
    conceptIds = c(1545958, 1549686, 1592085, 1539403)  # Atorvastatin, simvastatin, rosuvastatin, pravastatin
  ),
  
  # === MONITORING ===
  list(
    name = "Lipid Panel",
    domain = "Measurement",
    conceptIds = c(3027114, 3023103, 3019900, 3011884)  # Total chol, LDL, HDL, triglycerides
  ),
  list(
    name = "Cardiac Biomarkers",
    domain = "Measurement",
    conceptIds = c(3016407, 3005593, 3024561)  # Troponin, BNP, NT-proBNP
  )
)

# Create a cardiovascular cohort for testing
DatabaseConnector::executeSql(connection, "
  CREATE TABLE main.cv_cohort AS
  SELECT DISTINCT
    1 AS cohort_definition_id,
    co.person_id AS subject_id,
    MIN(co.condition_start_date) AS cohort_start_date,
    DATE(MIN(co.condition_start_date), '+1 year') AS cohort_end_date
  FROM main.condition_occurrence co
  WHERE co.condition_concept_id IN (4329847, 314666, 316139)  -- MI, angina, HF
  GROUP BY co.person_id
")

# Analyze with comprehensive filters
cvResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "cv_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = -30,
  endOffsetDays = 365,
  eventFilters = cvCarePathway,
  returnFormat = "list",
  verbose = TRUE
)

cli::cli_h3("ðŸ’” Cardiovascular Care Pathway Analysis")
print(cvResults$results)
```

### Dynamic Filter Generation

Create filters programmatically based on your needs:

```{r dynamic-filters}
# Function to create disease-specific filter sets
createDiseaseFilters <- function(diseaseName, conditionConceptIds, 
                                drugClasses, procedureTypes) {
  
  filters <- list()
  
  # Add condition filter
  filters[[length(filters) + 1]] <- list(
    name = paste(diseaseName, "Diagnoses"),
    domain = "Condition",
    conceptIds = conditionConceptIds
  )
  
  # Add drug filters for each class
  for(drugClass in names(drugClasses)) {
    filters[[length(filters) + 1]] <- list(
      name = paste(diseaseName, "-", drugClass),
      domain = "Drug",
      conceptIds = drugClasses[[drugClass]]
    )
  }
  
  # Add procedure filters
  for(procType in names(procedureTypes)) {
    filters[[length(filters) + 1]] <- list(
      name = paste(diseaseName, "-", procType),
      domain = "Procedure",
      conceptIds = procedureTypes[[procType]]
    )
  }
  
  return(filters)
}

# Example: Create diabetes filter set
diabetesDrugs <- list(
  "Metformin" = c(1503297, 1502826),
  "Sulfonylureas" = c(1502809, 1502855),
  "DPP4 Inhibitors" = c(1580747, 1583722),
  "GLP1 Agonists" = c(1583722, 44816332),
  "SGLT2 Inhibitors" = c(44785829, 45774435)
)

diabetesProcedures <- list(
  "Glucose Monitoring" = c(3004501, 3003309),
  "Eye Exams" = c(4079938, 4084167),
  "Foot Exams" = c(4030267, 4144723)
)

diabetesFilters <- createDiseaseFilters(
  "Diabetes",
  conditionConceptIds = c(201820, 201826, 443238, 442793),
  drugClasses = diabetesDrugs,
  procedureTypes = diabetesProcedures
)

# Use the generated filters
diabetesResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "cv_cohort",  # Using existing cohort for demo
  cohortId = 1,
  eventFilters = diabetesFilters,
  verbose = FALSE
)
```

## Multi-Cohort Analysis ðŸ‘¥

### Parallel Cohort Processing

Analyze multiple cohorts efficiently:

```{r multi-cohort}
# Create multiple cohorts for comparison
createComparisonCohorts <- function(connection) {
  
  # Drop existing table
  DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.comparison_cohorts")
  
  # Create cohort table
  DatabaseConnector::executeSql(connection, "
    CREATE TABLE main.comparison_cohorts (
      cohort_definition_id INTEGER,
      subject_id INTEGER,
      cohort_start_date DATE,
      cohort_end_date DATE
    )
  ")
  
  # Cohort 1: Diabetes only
  DatabaseConnector::executeSql(connection, "
    INSERT INTO main.comparison_cohorts
    SELECT 
      1 as cohort_definition_id,
      person_id as subject_id,
      MIN(condition_start_date) as cohort_start_date,
      DATE(MIN(condition_start_date), '+1 year') as cohort_end_date
    FROM main.condition_occurrence
    WHERE condition_concept_id IN (201820, 201826)
    GROUP BY person_id
  ")
  
  # Cohort 2: Hypertension only
  DatabaseConnector::executeSql(connection, "
    INSERT INTO main.comparison_cohorts
    SELECT 
      2 as cohort_definition_id,
      person_id as subject_id,
      MIN(condition_start_date) as cohort_start_date,
      DATE(MIN(condition_start_date), '+1 year') as cohort_end_date
    FROM main.condition_occurrence
    WHERE condition_concept_id IN (316866, 317895)
    GROUP BY person_id
  ")
  
  # Cohort 3: Both conditions
  DatabaseConnector::executeSql(connection, "
    INSERT INTO main.comparison_cohorts
    SELECT 
      3 as cohort_definition_id,
      a.subject_id,
      MAX(a.cohort_start_date, b.cohort_start_date) as cohort_start_date,
      MIN(a.cohort_end_date, b.cohort_end_date) as cohort_end_date
    FROM main.comparison_cohorts a
    JOIN main.comparison_cohorts b
      ON a.subject_id = b.subject_id
    WHERE a.cohort_definition_id = 1
      AND b.cohort_definition_id = 2
  ")
  
  # Get cohort counts
  counts <- DatabaseConnector::querySql(connection, "
    SELECT 
      cohort_definition_id,
      COUNT(DISTINCT subject_id) as n_patients
    FROM main.comparison_cohorts
    GROUP BY cohort_definition_id
  ")
  
  return(counts)
}

# Create cohorts
cohortCounts <- createComparisonCohorts(connection)
print(cohortCounts)

# Define cohort metadata
cohortMeta <- tribble(
  ~cohortId, ~cohortName, ~color,
  1, "Diabetes Only", "#e74c3c",
  2, "Hypertension Only", "#3498db",
  3, "Both Conditions", "#9b59b6"
)

# Analyze all cohorts with different time windows
timeWindows <- list(
  baseline = c(-365, -1),
  acute = c(0, 30),
  shortTerm = c(31, 90),
  longTerm = c(91, 365)
)

# Run analysis for all combinations
allResults <- expand_grid(
  cohortId = cohortMeta$cohortId,
  windowName = names(timeWindows)
) %>%
  pmap_dfr(function(cohortId, windowName) {
    window <- timeWindows[[windowName]]
    
    result <- calculateCostOfCare(
      connection = connection,
      cdmDatabaseSchema = "main",
      cohortDatabaseSchema = "main",
      cohortTable = "comparison_cohorts",
      cohortId = cohortId,
      anchorCol = "cohort_start_date",
      startOffsetDays = window[1],
      endOffsetDays = window[2],
      verbose = FALSE
    )
    
    result %>%
      mutate(
        cohortId = cohortId,
        windowName = windowName
      )
  }) %>%
  left_join(cohortMeta, by = "cohortId")

# Create comprehensive visualization
allResults %>%
  mutate(windowName = factor(windowName, levels = names(timeWindows))) %>%
  ggplot(aes(x = windowName, y = cost_pppm, color = cohortName, group = cohortName)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = setNames(cohortMeta$color, cohortMeta$cohortName)) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Cost Trajectories by Disease Combination",
    subtitle = "Comparing costs across different time windows relative to diagnosis",
    x = "Time Window",
    y = "Cost Per Person Per Month",
    color = "Cohort"
  ) +
  theme(legend.position = "bottom")
```

### Stratified Analysis

Perform analyses stratified by patient characteristics:

```{r stratified-analysis}
# Create age and gender stratified cohorts
createStratifiedCohorts <- function(connection) {
  
  DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.stratified_cohorts")
  
  # Create stratified cohorts with age groups and gender
  DatabaseConnector::executeSql(connection, "
    CREATE TABLE main.stratified_cohorts AS
    SELECT 
      c.*,
      p.gender_concept_id,
      CASE 
        WHEN (YEAR(c.cohort_start_date) - p.year_of_birth) < 40 THEN '18-39'
        WHEN (YEAR(c.cohort_start_date) - p.year_of_birth) < 65 THEN '40-64'
        ELSE '65+'
      END AS age_group
    FROM main.comparison_cohorts c
    JOIN main.person p ON c.subject_id = p.person_id
    WHERE c.cohort_definition_id = 1  -- Diabetes cohort
  ")
}

createStratifiedCohorts(connection)

# Define strata
strata <- expand_grid(
  age_group = c("18-39", "40-64", "65+"),
  gender = c(8507, 8532)  # Male, Female
) %>%
  mutate(
    stratum_id = row_number(),
    gender_name = if_else(gender == 8507, "Male", "Female")
  )

# Analyze each stratum
stratumResults <- pmap_dfr(strata, function(age_group, gender, stratum_id, gender_name) {
  
  # Create temporary cohort for this stratum
  sql <- sprintf("
    CREATE TEMPORARY TABLE temp_stratum AS
    SELECT * FROM main.stratified_cohorts
    WHERE age_group = '%s' AND gender_concept_id = %d
  ", age_group, gender)
  
  DatabaseConnector::executeSql(connection, sql)
  
  # Check if stratum has patients
  count <- DatabaseConnector::querySql(connection, 
    "SELECT COUNT(DISTINCT subject_id) as n FROM temp_stratum"
  )$n
  
  if(count > 0) {
    result <- calculateCostOfCare(
      connection = connection,
      cdmDatabaseSchema = "main",
      cohortDatabaseSchema = "main",
      cohortTable = "temp_stratum",
      cohortId = 1,
      anchorCol = "cohort_start_date",
      startOffsetDays = 0,
      endOffsetDays = 365,
      verbose = FALSE
    )
    
    result <- result %>%
      mutate(
        age_group = age_group,
        gender = gender_name,
        stratum_label = paste(age_group, gender_name, sep = ", ")
      )
  } else {
    result <- tibble(
      n_persons = 0,
      cost_pppm = 0,
      age_group = age_group,
      gender = gender_name,
      stratum_label = paste(age_group, gender_name, sep = ", ")
    )
  }
  
  # Clean up
  DatabaseConnector::executeSql(connection, "DROP TABLE temp_stratum")
  
  return(result)
})

# Visualize stratified results
stratumResults %>%
  filter(n_persons > 0) %>%
  ggplot(aes(x = age_group, y = cost_pppm, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Diabetes Costs by Age and Gender",
    subtitle = "First year following diagnosis",
    x = "Age Group",
    y = "Cost Per Person Per Month",
    fill = "Gender"
  )
```

## Performance Optimization âš¡

### Batch Processing Strategies

For large-scale analyses, use batch processing:

```{r batch-processing}
# Function for efficient batch processing
batchAnalyzeCohorts <- function(connection, cohortIds, batchSize = 10) {
  
  # Split cohorts into batches
  batches <- split(cohortIds, ceiling(seq_along(cohortIds) / batchSize))
  
  cli::cli_progress_bar("Processing cohort batches", total = length(batches))
  
  allResults <- map_dfr(batches, function(batch) {
    
    batchResults <- map_dfr(batch, function(cohortId) {
      
      result <- tryCatch({
        calculateCostOfCare(
          connection = connection,
          cdmDatabaseSchema = "main",
          cohortDatabaseSchema = "main",
          cohortTable = "comparison_cohorts",
          cohortId = cohortId,
          anchorCol = "cohort_start_date",
          startOffsetDays = 0,
          endOffsetDays = 365,
          verbose = FALSE
        ) %>%
        mutate(cohortId = cohortId)
      }, error = function(e) {
        cli::cli_alert_warning("Error processing cohort {cohortId}: {e$message}")
        tibble(cohortId = cohortId, error = TRUE)
      })
      
      return(result)
    })
    
    cli::cli_progress_update()
    return(batchResults)
  })
  
  cli::cli_progress_done()
  return(allResults)
}

# Example: Process multiple cohorts in batches
# results <- batchAnalyzeCohorts(connection, cohortIds = 1:100, batchSize = 10)
```

### Using Permanent Tables for Complex Analyses

For repeated analyses, use permanent helper tables:

```{r permanent-tables}
# Create analysis with permanent tables
runOptimizedAnalysis <- function(connection, cohortId, eventFilters) {
  
  # Generate unique table suffix
  tableSuffix <- format(Sys.time(), "%Y%m%d_%H%M%S")
  
  # Run analysis with permanent tables
  result <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "comparison_cohorts",
    cohortId = cohortId,
    anchorCol = "cohort_start_date",
    startOffsetDays = -365,
    endOffsetDays = 365,
    eventFilters = eventFilters,
    asPermanent = TRUE,  # Keep helper tables
    tableSuffix = tableSuffix,
    verbose = FALSE
  )
  
  # Store table names for cleanup
  helperTables <- c(
    paste0("cost_cohort_", tableSuffix),
    paste0("cost_events_", tableSuffix)
  )
  
  return(list(result = result, helperTables = helperTables))
}

# Example usage
# optimizedResult <- runOptimizedAnalysis(connection, 1, cvCarePathway)
```

## Advanced Diagnostics ðŸ“Š

### Detailed Patient Flow Analysis

Get comprehensive diagnostics about your analysis:

```{r diagnostics}
# Run analysis with full diagnostics
diagnosticResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "cv_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = -30,
  endOffsetDays = 365,
  eventFilters = cvCarePathway,
  returnFormat = "list",
  verbose = TRUE
)

# Analyze patient attrition
attrition <- diagnosticResults$diagnostics %>%
  mutate(
    step_order = row_number(),
    retention_rate = n_persons / first(n_persons) * 100,
    attrition_rate = 100 - retention_rate,
    persons_lost = lag(n_persons, default = first(n_persons)) - n_persons
  )

# Create detailed attrition plot
attrition %>%
  filter(step_name != "00_initial_cohort") %>%
  ggplot(aes(x = reorder(step_name, step_order))) +
  geom_bar(aes(y = n_persons), stat = "identity", fill = "#3498db", alpha = 0.7) +
  geom_line(aes(y = retention_rate * max(n_persons) / 100, group = 1), 
            color = "#e74c3c", size = 2) +
  geom_point(aes(y = retention_rate * max(n_persons) / 100), 
             color = "#c0392b", size = 3) +
  geom_text(aes(y = n_persons, label = paste0(round(retention_rate, 1), "%")), 
            vjust = -0.5, size = 3) +
  scale_y_continuous(
    name = "Number of Patients",
    sec.axis = sec_axis(~ . / max(attrition$n_persons) * 100, 
                        name = "Retention Rate (%)")
  ) +
  labs(
    title = "Patient Attrition Through Analysis Pipeline",
    subtitle = "Showing both absolute numbers and retention rates",
    x = "Analysis Step"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Event distribution analysis
if("n_events" %in% names(diagnosticResults$diagnostics)) {
  eventDist <- diagnosticResults$diagnostics %>%
    filter(n_events > 0) %>%
    mutate(
      events_per_person = n_events / n_persons,
      cost_per_event = if_else(n_events > 0, total_cost / n_events, 0)
    )
  
  # Create event analysis visualization
  eventDist %>%
    ggplot(aes(x = reorder(step_name, -events_per_person), 
               y = events_per_person)) +
    geom_bar(stat = "identity", fill = "#2ecc71") +
    geom_text(aes(label = round(events_per_person, 1)), vjust = -0.5) +
    labs(
      title = "Healthcare Utilization Intensity",
      subtitle = "Average events per person by category",
      x = "Event Category",
      y = "Events Per Person"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Custom Cost Concepts ðŸ’°

### Analyzing Different Cost Types

```{r cost-types}
# Define different cost concepts
costConcepts <- tribble(
  ~conceptId, ~conceptName, ~description,
  31978L, "Total Charge", "Full billed amount",
  31980L, "Total Cost", "Actual cost to payer",
  31981L, "Paid by Payer", "Amount paid by insurance",
  31982L, "Paid by Patient", "Out-of-pocket costs",
  31983L, "Paid by Primary", "Primary insurance payment",
  31984L, "Paid by Secondary", "Secondary insurance payment"
)

# Analyze each cost type
costTypeResults <- pmap_dfr(costConcepts, function(conceptId, conceptName, description) {
  
  result <- tryCatch({
    calculateCostOfCare(
      connection = connection,
      cdmDatabaseSchema = "main",
      cohortDatabaseSchema = "main",
      cohortTable = "cv_cohort",
      cohortId = 1,
      anchorCol = "cohort_start_date",
      startOffsetDays = 0,
      endOffsetDays = 365,
      costConceptId = conceptId,
      verbose = FALSE
    ) %>%
    mutate(
      costType = conceptName,
      description = description
    )
  }, error = function(e) {
    tibble(
      costType = conceptName,
      total_cost = NA_real_,
      cost_pppm = NA_real_,
      description = description,
      error = TRUE
    )
  })
  
  return(result)
})

# Visualize cost breakdown
costTypeResults %>%
  filter(!is.na(total_cost)) %>%
  ggplot(aes(x = reorder(costType, total_cost), y = total_cost)) +
  geom_bar(stat = "identity", fill = "#34495e") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Healthcare Costs by Payment Type",
    subtitle = "Different perspectives on the same care",
    x = "Cost Type",
    y = "Total Cost"
  )
```

## Time Window Optimization ðŸ“…

### Finding Optimal Analysis Windows

```{r window-optimization}
# Test different window sizes to find optimal analysis period
windowSizes <- c(7, 14, 30, 60, 90, 180, 365)

windowAnalysis <- map_dfr(windowSizes, function(days) {
  
  result <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "cv_cohort",
    cohortId = 1,
    anchorCol = "cohort_start_date",
    startOffsetDays = 0,
    endOffsetDays = days,
    verbose = FALSE
  )
  
  result %>%
    mutate(
      windowDays = days,
      captureRate = n_persons / first(n_persons) * 100
    )
})

# Visualize capture rates
windowAnalysis %>%
  ggplot(aes(x = windowDays)) +
  geom_line(aes(y = captureRate), color = "#3498db", size = 1.5) +
  geom_point(aes(y = captureRate), color = "#2980b9", size = 3) +
  geom_hline(yintercept = c(80, 90, 95), linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = windowSizes) +
  labs(
    title = "Patient Capture Rate by Analysis Window",
    subtitle = "Percentage of cohort with cost data by window size",
    x = "Window Size (days)",
    y = "Capture Rate (%)"
  )
```

## Clean Up

```{r cleanup}
# Remove all temporary tables
tables_to_clean <- c(
  "cv_cohort", "comparison_cohorts", "stratified_cohorts"
)

for(table in tables_to_clean) {
  DatabaseConnector::executeSql(
    connection, 
    sprintf("DROP TABLE IF EXISTS main.%s", table)
  )
}

# Disconnect
DatabaseConnector::disconnect(connection)
cli::cli_alert_success("Advanced analysis complete! ðŸŽ‰")
```

## Key Takeaways ðŸŽ¯

You've mastered:

1. **Complex Event Filtering** - Build comprehensive clinical pathways
2. **Multi-Cohort Analysis** - Compare multiple populations efficiently
3. **Stratified Analysis** - Examine subgroups within cohorts
4. **Performance Optimization** - Handle large-scale analyses
5. **Advanced Diagnostics** - Understand patient flow in detail
6. **Custom Cost Concepts** - Analyze different payment perspectives

## Next Steps

- Combine these techniques for your specific research questions
- Explore the package source code for additional options
- Share your analyses with the OHDSI community
- Contribute improvements back to the package

Remember: With great power comes great responsibility. Always validate your results and consider the clinical context!

Happy analyzing! ðŸš€