---
title: "Adjusting Healthcare Costs for Inflation Using CPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjusting Healthcare Costs for Inflation Using CPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Why Adjust for Inflation? ðŸ’¸

Healthcare costs change over time, but how much of that change is real versus just inflation? When analyzing costs across multiple years, adjusting for inflation helps you:

- ðŸ“Š **Make fair comparisons** between different time periods
- ðŸ“ˆ **Identify real trends** versus inflationary growth
- ðŸŽ¯ **Standardize results** to a common year for reporting
- ðŸ’¡ **Understand true cost drivers** in healthcare

This vignette shows you how to use the CostUtilization package's built-in CPI (Consumer Price Index) adjustment features.

```{r setup}
library(CostUtilization)
library(DatabaseConnector)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(cli)

# Set theme for nice plots
theme_set(theme_minimal() + 
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(color = "gray50")))
```

## Understanding CPI Basics ðŸ“š

The Consumer Price Index (CPI) tracks how prices change over time. For healthcare:

- **Medical Care CPI**: Specific to healthcare services
- **All Items CPI**: General inflation across all goods/services
- **Base Year**: Reference point (often 100)
- **Adjustment Factor**: Multiplier to convert between years

### Loading and Exploring CPI Data

```{r explore-cpi}
# Load the default medical care CPI data
cpiData <- loadCpiData()

# Examine the data
cli::cli_h3("ðŸ“Š CPI Data Structure")
head(cpiData, 10)

# Calculate average annual inflation
cpiData <- cpiData %>%
  arrange(year) %>%
  mutate(
    annual_inflation = (cpi / lag(cpi) - 1) * 100
  )

# Summary statistics
avgInflation <- mean(cpiData$annual_inflation, na.rm = TRUE)
cli::cli_alert_info("Average annual medical inflation: {round(avgInflation, 2)}%")

# Visualize CPI trend
ggplot(cpiData, aes(x = year)) +
  geom_line(aes(y = cpi), color = "#3498db", size = 1.5) +
  geom_point(aes(y = cpi), color = "#2980b9", size = 2) +
  geom_smooth(aes(y = cpi), method = "loess", se = TRUE, alpha = 0.2) +
  labs(
    title = "Medical Care CPI Over Time",
    subtitle = "Rising healthcare costs reflected in CPI values",
    x = "Year",
    y = "CPI Index Value",
    caption = "Source: U.S. Bureau of Labor Statistics"
  )

# Show inflation rate over time
cpiData %>%
  filter(!is.na(annual_inflation)) %>%
  ggplot(aes(x = year, y = annual_inflation)) +
  geom_col(fill = "#e74c3c", alpha = 0.7) +
  geom_hline(yintercept = avgInflation, linetype = "dashed", color = "#c0392b") +
  geom_text(aes(label = round(annual_inflation, 1)), vjust = -0.5, size = 3) +
  labs(
    title = "Annual Medical Care Inflation Rate",
    subtitle = paste("Average:", round(avgInflation, 2), "% per year"),
    x = "Year",
    y = "Annual Inflation Rate (%)"
  )
```

## Basic CPI Calculations ðŸ§®

### Understanding Adjustment Factors

```{r basic-calculations}
# Calculate adjustment factors between years
calculateAndVisualizeCPI <- function(fromYear, toYear, cpiData) {
  
  # Get CPI values
  fromCPI <- cpiData$cpi[cpiData$year == fromYear]
  toCPI <- cpiData$cpi[cpiData$year == toYear]
  
  # Calculate adjustment factor
  adjustmentFactor <- toCPI / fromCPI
  
  # Create example costs
  exampleCosts <- c(1000, 5000, 10000, 50000, 100000)
  
  # Calculate adjusted values
  adjustedCosts <- exampleCosts * adjustmentFactor
  
  # Create comparison table
  comparison <- tibble(
    original_cost = exampleCosts,
    year = fromYear,
    adjusted_cost = adjustedCosts,
    adjusted_year = toYear,
    percent_increase = (adjustedCosts - exampleCosts) / exampleCosts * 100
  )
  
  cli::cli_h3("ðŸ’° Cost Adjustment from {fromYear} to {toYear}")
  cli::cli_alert_info("Adjustment factor: {round(adjustmentFactor, 3)}")
  cli::cli_alert_info("Total inflation: {round((adjustmentFactor - 1) * 100, 1)}%")
  
  print(comparison)
  
  # Visualize the adjustment
  comparison %>%
    pivot_longer(cols = c(original_cost, adjusted_cost), 
                 names_to = "type", 
                 values_to = "cost") %>%
    mutate(
      type = if_else(type == "original_cost", 
                     paste(fromYear, "Dollars"), 
                     paste(toYear, "Dollars"))
    ) %>%
    ggplot(aes(x = factor(original_cost), y = cost, fill = type)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_y_continuous(labels = dollar) +
    scale_fill_manual(values = c("#3498db", "#e74c3c")) +
    labs(
      title = "Impact of Medical Inflation on Healthcare Costs",
      subtitle = sprintf("Converting %d dollars to %d dollars", fromYear, toYear),
      x = "Original Cost (2015 $)",
      y = "Cost Value",
      fill = "Year"
    ) +
    theme(legend.position = "bottom")
}

# Example: 2015 to 2023 adjustment
calculateAndVisualizeCPI(2015, 2023, cpiData)

# Example: 2010 to 2023 adjustment (longer period)
calculateAndVisualizeCPI(2010, 2023, cpiData)
```

## Using CPI in Cost Analysis ðŸ“Š

### Setting Up Test Data

```{r setup-analysis}
# Connect to database (using Eunomia for example)
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
connection <- DatabaseConnector::connect(connectionDetails)

# Add cost data
injectCostData(connection, cdmDatabaseSchema = "main", seed = 123)
transformCostToCdmV5dot4(connectionDetails, cdmDatabaseSchema = "main")

# Create a multi-year cohort for demonstration
DatabaseConnector::executeSql(connection, "
  CREATE TABLE main.multiyear_cohort AS
  SELECT 
    1 AS cohort_definition_id,
    person_id AS subject_id,
    visit_start_date AS cohort_start_date,
    DATE(visit_start_date, '+1 year') AS cohort_end_date
  FROM main.visit_occurrence
  WHERE YEAR(visit_start_date) BETWEEN 2010 AND 2019
    AND visit_concept_id = 9201  -- Inpatient visits
")

cohortInfo <- DatabaseConnector::querySql(connection, "
  SELECT 
    YEAR(cohort_start_date) as year,
    COUNT(DISTINCT subject_id) as n_patients
  FROM main.multiyear_cohort
  GROUP BY YEAR(cohort_start_date)
  ORDER BY year
")

cli::cli_h3("ðŸ“… Multi-year Cohort Distribution")
print(cohortInfo)
```

### Basic CPI-Adjusted Analysis

```{r basic-adjusted-analysis}
# Run analysis WITHOUT CPI adjustment
unadjustedResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "multiyear_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = 0,
  endOffsetDays = 365,
  cpiAdjustment = FALSE,
  verbose = TRUE
)

# Run analysis WITH CPI adjustment to 2023 dollars
adjustedResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "multiyear_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = 0,
  endOffsetDays = 365,
  cpiAdjustment = TRUE,
  cpiTargetYear = 2023,
  returnFormat = "list",
  verbose = TRUE
)

# Compare results
comparison <- bind_rows(
  unadjustedResults %>% 
    select(n_persons, total_cost, cost_pppm) %>%
    mutate(type = "Nominal (Unadjusted)"),
  adjustedResults$results %>% 
    select(n_persons, total_cost = adjusted_cost, cost_pppm = adjusted_cost_pppm) %>%
    mutate(type = "Adjusted to 2023 $")
)

cli::cli_h3("ðŸ’µ Comparison: Adjusted vs Unadjusted Costs")
print(comparison)

# Calculate the impact
impact <- (comparison$total_cost[2] - comparison$total_cost[1]) / comparison$total_cost[1] * 100
cli::cli_alert_info("CPI adjustment increased total costs by {round(impact, 1)}%")
```

## Multi-Year Trend Analysis ðŸ“ˆ

### Analyzing Cost Trends Over Time

```{r trend-analysis}
# Create year-specific cohorts
years <- 2010:2019

# Create cohorts for each year
for(year in years) {
  sql <- sprintf("
    CREATE TABLE main.cohort_%d AS
    SELECT 
      1 AS cohort_definition_id,
      person_id AS subject_id,
      MIN(visit_start_date) AS cohort_start_date,
      DATE(MIN(visit_start_date), '+1 year') AS cohort_end_date
    FROM main.visit_occurrence
    WHERE YEAR(visit_start_date) = %d
      AND visit_concept_id IN (9201, 9203)  -- Inpatient and ER
    GROUP BY person_id
  ", year, year)
  
  DatabaseConnector::executeSql(connection, sql)
}

# Analyze each year with and without adjustment
yearlyResults <- map_dfr(years, function(year) {
  
  # Unadjusted analysis
  unadjusted <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = paste0("cohort_", year),
    cohortId = 1,
    anchorCol = "cohort_start_date",
    startOffsetDays = 0,
    endOffsetDays = 365,
    cpiAdjustment = FALSE,
    verbose = FALSE
  )
  
  # Adjusted analysis
  adjusted <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = paste0("cohort_", year),
    cohortId = 1,
    anchorCol = "cohort_start_date",
    startOffsetDays = 0,
    endOffsetDays = 365,
    cpiAdjustment = TRUE,
    cpiTargetYear = 2023,
    verbose = FALSE
  )
  
  # Combine results
  tibble(
    year = year,
    n_patients = unadjusted$n_persons,
    nominal_cost = unadjusted$cost_pppm,
    adjusted_cost = adjusted$adjusted_cost_pppm,
    nominal_total = unadjusted$total_cost,
    adjusted_total = adjusted$adjusted_cost
  )
})

# Create trend visualization
yearlyResults %>%
  pivot_longer(cols = c(nominal_cost, adjusted_cost), 
               names_to = "type", 
               values_to = "cost_pppm") %>%
  mutate(
    type = if_else(type == "nominal_cost", "Nominal Dollars", "2023 Dollars")
  ) %>%
  ggplot(aes(x = year, y = cost_pppm, color = type)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Nominal Dollars" = "#3498db", 
                                "2023 Dollars" = "#e74c3c")) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Healthcare Cost Trends: Nominal vs Inflation-Adjusted",
    subtitle = "Cost per person per month for hospital-based care",
    x = "Year",
    y = "Cost PPPM",
    color = "Cost Type",
    caption = "Adjusted using medical care CPI to 2023 dollars"
  ) +
  theme(legend.position = "bottom")

# Calculate real vs nominal growth
firstYear <- yearlyResults$year[1]
lastYear <- yearlyResults$year[nrow(yearlyResults)]
nominalGrowth <- (yearlyResults$nominal_cost[nrow(yearlyResults)] / 
                  yearlyResults$nominal_cost[1] - 1) * 100
realGrowth <- (yearlyResults$adjusted_cost[nrow(yearlyResults)] / 
               yearlyResults$adjusted_cost[1] - 1) * 100

cli::cli_h3("ðŸ“Š Growth Analysis ({firstYear} to {lastYear})")
cli::cli_alert_info("Nominal growth: {round(nominalGrowth, 1)}%")
cli::cli_alert_info("Real growth (inflation-adjusted): {round(realGrowth, 1)}%")
cli::cli_alert_info("Inflation impact: {round(nominalGrowth - realGrowth, 1)} percentage points")
```

## Using Custom CPI Data ðŸ› ï¸

### Creating and Using Custom CPI Tables

```{r custom-cpi}
# Example: Create custom CPI data for a specific region or service
customCpiData <- tibble(
  year = 2010:2023,
  cpi = c(100, 103, 106, 109, 111, 114, 117, 120, 
          124, 128, 132, 137, 143, 150)  # Example values
)

# Add calculated inflation rates
customCpiData <- customCpiData %>%
  mutate(
    annual_inflation = (cpi / lag(cpi) - 1) * 100,
    cumulative_inflation = (cpi / first(cpi) - 1) * 100
  )

# Visualize custom CPI
customCpiData %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = cumulative_inflation), color = "#9b59b6", size = 1.5) +
  geom_point(aes(y = cumulative_inflation), color = "#8e44ad", size = 3) +
  geom_area(aes(y = cumulative_inflation), fill = "#9b59b6", alpha = 0.2) +
  labs(
    title = "Custom CPI: Cumulative Inflation Over Time",
    subtitle = "Example of region-specific healthcare inflation",
    x = "Year",
    y = "Cumulative Inflation (%)",
    caption = "Base year: 2010 = 100"
  )

# Save custom CPI data
write.csv(customCpiData, "custom_healthcare_cpi.csv", row.names = FALSE)

# Use custom CPI in analysis
customAdjustedResults <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "multiyear_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = 0,
  endOffsetDays = 365,
  cpiAdjustment = TRUE,
  cpiTargetYear = 2023,
  cpiDataPath = "custom_healthcare_cpi.csv",
  verbose = FALSE
)

# Compare different CPI adjustments
cpiComparison <- bind_rows(
  unadjustedResults %>% 
    mutate(adjustment = "None"),
  adjustedResults$results %>% 
    select(total_cost = adjusted_cost, cost_pppm = adjusted_cost_pppm) %>%
    mutate(adjustment = "Standard Medical CPI"),
  customAdjustedResults %>% 
    select(total_cost = adjusted_cost, cost_pppm = adjusted_cost_pppm) %>%
    mutate(adjustment = "Custom Regional CPI")
)

# Visualize comparison
cpiComparison %>%
  ggplot(aes(x = adjustment, y = cost_pppm, fill = adjustment)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("None" = "#95a5a6", 
                               "Standard Medical CPI" = "#3498db",
                               "Custom Regional CPI" = "#9b59b6")) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Impact of Different CPI Adjustments",
    subtitle = "Same cohort analyzed with different inflation adjustments",
    x = "Adjustment Type",
    y = "Cost Per Person Per Month"
  ) +
  theme(legend.position = "none")
```

## Advanced CPI Features ðŸš€

### Cohort-Specific CPI Adjustment

```{r advanced-cpi}
# Function to apply cohort-specific CPI adjustments
applyCohortSpecificCPI <- function(connection, cohortTable, targetYear = 2023) {
  
  # Get cohort years
  cohortYears <- DatabaseConnector::querySql(connection, sprintf("
    SELECT DISTINCT YEAR(cohort_start_date) as cohort_year
    FROM main.%s
    ORDER BY cohort_year
  ", cohortTable))
  
  # Analyze each cohort year separately
  results <- map_dfr(cohortYears$COHORT_YEAR, function(year) {
    
    # Create year-specific temp table
    DatabaseConnector::executeSql(connection, sprintf("
      CREATE TEMPORARY TABLE temp_cohort_%d AS
      SELECT * FROM main.%s
      WHERE YEAR(cohort_start_date) = %d
    ", year, cohortTable, year))
    
    # Run analysis with appropriate CPI adjustment
    result <- calculateCostOfCare(
      connection = connection,
      cdmDatabaseSchema = "main",
      cohortDatabaseSchema = "main",
      cohortTable = sprintf("temp_cohort_%d", year),
      cohortId = 1,
      anchorCol = "cohort_start_date",
      startOffsetDays = 0,
      endOffsetDays = 365,
      cpiAdjustment = TRUE,
      cpiTargetYear = targetYear,
      cpiFromYear = year,  # Specify the source year
      verbose = FALSE
    )
    
    # Clean up
    DatabaseConnector::executeSql(connection, 
      sprintf("DROP TABLE temp_cohort_%d", year))
    
    result %>%
      mutate(cohort_year = year)
  })
  
  return(results)
}

# Example usage
# cohortSpecificResults <- applyCohortSpecificCPI(connection, "multiyear_cohort")
```

### Creating CPI Adjustment Reports

```{r cpi-reports}
# Function to create comprehensive CPI adjustment report
createCPIAdjustmentReport <- function(results, cpiData, targetYear) {
  
  # Calculate summary statistics
  summary_stats <- results %>%
    summarise(
      total_patients = sum(n_patients),
      total_nominal_cost = sum(nominal_total),
      total_adjusted_cost = sum(adjusted_total),
      avg_nominal_pppm = mean(nominal_cost),
      avg_adjusted_pppm = mean(adjusted_cost),
      inflation_impact = (total_adjusted_cost - total_nominal_cost) / total_nominal_cost * 100
    )
  
  # Create report visualization
  p1 <- results %>%
    ggplot(aes(x = year)) +
    geom_bar(aes(y = n_patients), stat = "identity", fill = "#34495e", alpha = 0.7) +
    labs(title = "Patient Distribution by Year", y = "Number of Patients")
  
  p2 <- results %>%
    pivot_longer(cols = c(nominal_total, adjusted_total), 
                 names_to = "type", values_to = "cost") %>%
    mutate(type = if_else(type == "nominal_total", "Nominal", "Adjusted")) %>%
    ggplot(aes(x = year, y = cost, fill = type)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_y_continuous(labels = dollar) +
    scale_fill_manual(values = c("Nominal" = "#3498db", "Adjusted" = "#e74c3c")) +
    labs(title = "Total Costs by Year", y = "Total Cost", fill = "Type")
  
  # Print report
  cli::cli_h2("ðŸ“Š CPI Adjustment Report")
  cli::cli_h3("Summary Statistics")
  cli::cli_alert_info("Target year for adjustment: {targetYear}")
  cli::cli_alert_info("Total patients analyzed: {comma(summary_stats$total_patients)}")
  cli::cli_alert_info("Total nominal cost: {dollar(summary_stats$total_nominal_cost)}")
  cli::cli_alert_info("Total adjusted cost: {dollar(summary_stats$total_adjusted_cost)}")
  cli::cli_alert_info("Overall inflation impact: {round(summary_stats$inflation_impact, 1)}%")
  
  return(list(
    summary = summary_stats,
    patient_plot = p1,
    cost_plot = p2
  ))
}

# Generate report
report <- createCPIAdjustmentReport(yearlyResults, cpiData, 2023)
print(report$patient_plot)
print(report$cost_plot)
```

## Best Practices and Tips ðŸ’¡

### 1. Choose the Right CPI

```{r cpi-selection}
# Compare different CPI types
cpiTypes <- tribble(
  ~type, ~description, ~use_case,
  "Medical Care", "Healthcare-specific inflation", "Most healthcare cost analyses",
  "Medical Services", "Professional services only", "Physician cost studies",
  "Hospital Services", "Inpatient care inflation", "Hospital-specific analyses",
  "Prescription Drugs", "Pharmaceutical inflation", "Drug cost studies",
  "All Items", "General inflation", "Comparison with overall economy"
)

cli::cli_h3("ðŸ“‹ CPI Selection Guide")
print(cpiTypes)
```

### 2. Document Your Adjustments

Always document:
- Which CPI series you used
- The target year for adjustment
- Any custom modifications
- Rationale for your choices

### 3. Consider Regional Variations

Healthcare inflation varies by region. Consider using:
- Regional CPI data when available
- State-specific adjustments for local studies
- Urban vs rural adjustments

### 4. Validate Your Results

```{r validation}
# Example validation checks
validateCPIAdjustment <- function(unadjusted, adjusted, cpiData, fromYear, toYear) {
  
  # Calculate expected adjustment
  expectedFactor <- cpiData$cpi[cpiData$year == toYear] / 
                   cpiData$cpi[cpiData$year == fromYear]
  
  # Calculate actual adjustment
  actualFactor <- adjusted$adjusted_cost / unadjusted$total_cost
  
  # Check if adjustment is reasonable
  difference <- abs(actualFactor - expectedFactor) / expectedFactor * 100
  
  cli::cli_h3("âœ… CPI Adjustment Validation")
  cli::cli_alert_info("Expected adjustment factor: {round(expectedFactor, 3)}")
  cli::cli_alert_info("Actual adjustment factor: {round(actualFactor, 3)}")
  
  if(difference < 5) {
    cli::cli_alert_success("Adjustment validated successfully (difference: {round(difference, 1)}%)")
  } else {
    cli::cli_alert_warning("Large difference detected: {round(difference, 1)}%")
  }
  
  return(difference < 5)
}
```

## Clean Up

```{r cleanup}
# Remove temporary tables
for(year in years) {
  DatabaseConnector::executeSql(connection, 
    sprintf("DROP TABLE IF EXISTS main.cohort_%d", year))
}

DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.multiyear_cohort")

# Remove custom CPI file
if(file.exists("custom_healthcare_cpi.csv")) {
  file.remove("custom_healthcare_cpi.csv")
}

# Disconnect
DatabaseConnector::disconnect(connection)
cli::cli_alert_success("Clean up complete!")
```

## Summary ðŸŽ¯

You've learned how to:

- âœ… Understand CPI concepts and calculations
- âœ… Apply CPI adjustments to cost analyses
- âœ… Analyze multi-year trends with inflation adjustment
- âœ… Use custom CPI data for specific needs
- âœ… Create comprehensive CPI adjustment reports
- âœ… Validate and document your adjustments

Remember: CPI adjustment is a powerful tool for making healthcare costs comparable across time, but always consider the context and document your methodology clearly.

## Next Steps

1. Explore different CPI series for your specific use case
2. Consider creating region-specific CPI tables
3. Combine CPI adjustment with other advanced features
4. Share your inflation-adjusted analyses with stakeholders

Happy analyzing! ðŸ“Š