---
title: "Working with Eunomia Test Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Eunomia Test Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Introduction to Eunomia ðŸŒŸ

Eunomia is OHDSI's gift to developers - a synthetic dataset that lets you test and learn without worrying about patient privacy. This vignette shows you how to make the most of Eunomia for cost analysis development and testing.

### Why Use Eunomia?

- ðŸ”’ **No privacy concerns** - 100% synthetic data
- ðŸš€ **Quick setup** - No database installation needed
- ðŸ”„ **Reproducible** - Same data every time
- ðŸ“š **Educational** - Perfect for learning and demos

## Setting Up Your Environment

```{r setup}
# Load required packages
library(CostUtilization)
library(DatabaseConnector)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cli)

# Install Eunomia if needed
if (!requireNamespace("Eunomia", quietly = TRUE)) {
  cli::cli_alert_info("Installing Eunomia package...")
  install.packages("Eunomia")
}

library(Eunomia)

# Set a theme for nice plots
theme_set(theme_minimal())
```

## Understanding Eunomia's Structure

Let's explore what's in the Eunomia database:

```{r explore-eunomia}
# Connect to Eunomia
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
connection <- DatabaseConnector::connect(connectionDetails)

cli::cli_alert_success("Connected to Eunomia!")

# List all tables
tables <- DatabaseConnector::getTableNames(connection, "main")
cli::cli_h3("ðŸ“‹ Available tables in Eunomia:")
print(sort(tables))

# Check key statistics
stats <- DatabaseConnector::querySql(connection, "
  SELECT 
    (SELECT COUNT(DISTINCT person_id) FROM main.person) as n_persons,
    (SELECT COUNT(*) FROM main.visit_occurrence) as n_visits,
    (SELECT COUNT(*) FROM main.condition_occurrence) as n_conditions,
    (SELECT COUNT(*) FROM main.drug_exposure) as n_drugs,
    (SELECT COUNT(*) FROM main.procedure_occurrence) as n_procedures
")

cli::cli_h3("ðŸ“Š Database Statistics:")
print(stats)
```

## Adding Synthetic Cost Data

Eunomia doesn't include cost data by default, but we can add realistic synthetic costs:

```{r inject-costs}
cli::cli_h2("ðŸ’° Creating Synthetic Cost Data")

# Inject cost data with custom parameters
injectCostData(
  connection = connection,
  cdmDatabaseSchema = "main",
  costDomains = c("Procedure", "Drug", "Visit", "Device", "Measurement", "Observation", "Condition"),
  costTypes = c("Charge", "Paid", "Copay"),  # Multiple cost types
  seed = 123  # For reproducibility
)

# Check what was created
costStats <- DatabaseConnector::querySql(connection, "
  SELECT 
    cost_domain_id,
    COUNT(*) as n_records,
    AVG(cost) as avg_cost,
    MIN(cost) as min_cost,
    MAX(cost) as max_cost
  FROM main.cost
  WHERE cost_concept_id = 31978  -- Total charge
  GROUP BY cost_domain_id
")

cli::cli_h3("ðŸ’µ Cost Data Summary by Domain:")
print(costStats)

# Also check payer plan periods
payerStats <- DatabaseConnector::querySql(connection, "
  SELECT 
    payer_source_value,
    COUNT(*) as n_periods,
    COUNT(DISTINCT person_id) as n_persons
  FROM main.payer_plan_period
  GROUP BY payer_source_value
")

cli::cli_h3("ðŸ¥ Payer Plan Distribution:")
print(payerStats)
```

## Transforming to Modern CDM Format

```{r transform-costs}
cli::cli_h2("ðŸ”„ Transforming Cost Table to CDM v5.4+ Format")

# Transform the cost table
transformCostToCdmV5dot4(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  createIndexes = TRUE  # Important for performance
)

# Verify the transformation
costStructure <- DatabaseConnector::querySql(connection, "
  SELECT 
    cost_concept_id,
    COUNT(*) as n_records,
    COUNT(DISTINCT cost_event_id) as n_unique_events
  FROM main.cost
  GROUP BY cost_concept_id
")

cli::cli_h3("âœ… Transformed Cost Table Structure:")
print(costStructure)
```

## Example Analyses with Eunomia

### Example 1: Analyzing Costs for Common Conditions

Let's analyze costs for some common conditions in Eunomia:

```{r common-conditions}
# Find the most common conditions
commonConditions <- DatabaseConnector::querySql(connection, "
  SELECT 
    c.concept_name,
    c.concept_id,
    COUNT(*) as n_occurrences,
    COUNT(DISTINCT co.person_id) as n_patients
  FROM main.condition_occurrence co
  JOIN main.concept c ON co.condition_concept_id = c.concept_id
  GROUP BY c.concept_id, c.concept_name
  ORDER BY n_patients DESC
  LIMIT 10
")

cli::cli_h3("ðŸ¥ Most Common Conditions in Eunomia:")
print(commonConditions)

# Create cohorts for top 3 conditions
topConditions <- commonConditions[1:3,]

# Create cohort table
DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.condition_cohorts")
DatabaseConnector::executeSql(connection, "
  CREATE TABLE main.condition_cohorts (
    cohort_definition_id INTEGER,
    subject_id INTEGER,
    cohort_start_date DATE,
    cohort_end_date DATE
  )
")

# Populate cohorts
for(i in 1:nrow(topConditions)) {
  sql <- sprintf("
    INSERT INTO main.condition_cohorts
    SELECT 
      %d as cohort_definition_id,
      person_id as subject_id,
      MIN(condition_start_date) as cohort_start_date,
      DATE(MIN(condition_start_date), '+90 days') as cohort_end_date
    FROM main.condition_occurrence
    WHERE condition_concept_id = %d
    GROUP BY person_id
  ", i, topConditions$CONCEPT_ID[i])
  
  DatabaseConnector::executeSql(connection, sql)
}

# Analyze costs for each condition
conditionResults <- map_dfr(1:3, function(cohortId) {
  result <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "condition_cohorts",
    cohortId = cohortId,
    anchorCol = "cohort_start_date",
    startOffsetDays = 0,
    endOffsetDays = 90,
    verbose = FALSE
  )
  
  result %>%
    mutate(
      condition = topConditions$CONCEPT_NAME[cohortId],
      cohortId = cohortId
    )
})

# Visualize results
ggplot(conditionResults, aes(x = reorder(condition, -cost_pppm), y = cost_pppm)) +
  geom_bar(stat = "identity", fill = "#3498db") +
  geom_text(aes(label = paste0("$", round(cost_pppm, 0))), vjust = -0.5) +
  labs(
    title = "Cost Comparison Across Common Conditions",
    subtitle = "90-day costs following first diagnosis",
    x = "Condition",
    y = "Cost Per Person Per Month ($)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Example 2: Drug Cost Comparison Study

Let's compare costs between different drug exposures:

```{r drug-comparison}
# Common drugs in Eunomia for comparison
drugCohorts <- tribble(
  ~cohortId, ~drugConceptId, ~drugName,
  1, 1118084, "Celecoxib",
  2, 1124300, "Diclofenac",
  3, 1125315, "Acetaminophen"
)

# Create drug exposure cohorts
DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.drug_cohorts")

createDrugCohortSql <- "
  CREATE TABLE main.drug_cohorts AS
  SELECT 
    1 as cohort_definition_id,
    person_id as subject_id,
    MIN(drug_exposure_start_date) as cohort_start_date,
    DATE(MIN(drug_exposure_start_date), '+30 days') as cohort_end_date
  FROM main.drug_exposure
  WHERE drug_concept_id = 1118084
  GROUP BY person_id
"

DatabaseConnector::executeSql(connection, createDrugCohortSql)

# Add other drug cohorts
for(i in 2:nrow(drugCohorts)) {
  sql <- sprintf("
    INSERT INTO main.drug_cohorts
    SELECT 
      %d as cohort_definition_id,
      person_id as subject_id,
      MIN(drug_exposure_start_date) as cohort_start_date,
      DATE(MIN(drug_exposure_start_date), '+30 days') as cohort_end_date
    FROM main.drug_exposure
    WHERE drug_concept_id = %d
    GROUP BY person_id
  ", i, drugCohorts$drugConceptId[i])
  
  DatabaseConnector::executeSql(connection, sql)
}

# Analyze with event filters for related conditions
painFilters <- list(
  list(
    name = "Pain Diagnoses",
    domain = "Condition",
    conceptIds = c(4129418, 77670, 4144583)  # Various pain conditions
  ),
  list(
    name = "Pain Procedures",
    domain = "Procedure",
    conceptIds = c(4142875, 4298121)  # Pain-related procedures
  )
)

# Run analysis for each drug
drugResults <- map_dfr(1:nrow(drugCohorts), function(cohortId) {
  result <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "drug_cohorts",
    cohortId = cohortId,
    anchorCol = "cohort_start_date",
    startOffsetDays = -7,  # Week before
    endOffsetDays = 30,    # Month after
    eventFilters = painFilters,
    verbose = FALSE
  )
  
  result %>%
    mutate(drug = drugCohorts$drugName[cohortId])
})

# Create comparison visualization
drugResults %>%
  pivot_longer(cols = c(total_cost, n_visits), names_to = "metric", values_to = "value") %>%
  mutate(
    metric_label = case_when(
      metric == "total_cost" ~ "Total Cost ($)",
      metric == "n_visits" ~ "Number of Visits"
    )
  ) %>%
  ggplot(aes(x = drug, y = value, fill = drug)) +
  geom_bar(stat = "identity") +
  facet_wrap(~metric_label, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Pain Medication Cost and Utilization Comparison",
    subtitle = "Costs and visits from 1 week before to 1 month after initiation",
    x = "Medication",
    y = "Value"
  ) +
  theme(legend.position = "none")
```

### Example 3: Time Series Analysis

Let's look at cost patterns over time:

```{r time-series}
# Create a cohort with clear index date (first GI bleed)
DatabaseConnector::executeSql(connection, "DROP TABLE IF EXISTS main.gi_bleed_cohort")
DatabaseConnector::executeSql(connection, "
  CREATE TABLE main.gi_bleed_cohort AS
  SELECT DISTINCT
    1 AS cohort_definition_id,
    person_id AS subject_id,
    MIN(condition_start_date) AS cohort_start_date,
    DATE(MIN(condition_start_date), '+1 year') AS cohort_end_date
  FROM main.condition_occurrence
  WHERE condition_concept_id IN (192671, 4112343)
  GROUP BY person_id
")

# Define monthly windows for a year
monthlyWindows <- tibble(
  month = 1:12,
  startDays = (0:11) * 30,
  endDays = (1:12) * 30 - 1
)

# Analyze each month
monthlyResults <- pmap_dfr(monthlyWindows, function(month, startDays, endDays) {
  result <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "gi_bleed_cohort",
    cohortId = 1,
    anchorCol = "cohort_start_date",
    startOffsetDays = startDays,
    endOffsetDays = endDays,
    verbose = FALSE
  )
  
  result %>%
    mutate(
      month = month,
      monthLabel = paste("Month", month)
    )
})

# Create time series visualization
ggplot(monthlyResults, aes(x = month)) +
  geom_line(aes(y = cost_pppm), color = "#e74c3c", size = 1.5) +
  geom_point(aes(y = cost_pppm), color = "#c0392b", size = 3) +
  geom_area(aes(y = cost_pppm), fill = "#e74c3c", alpha = 0.2) +
  scale_x_continuous(breaks = 1:12, labels = paste("M", 1:12)) +
  labs(
    title = "Monthly Cost Trajectory Following GI Bleed",
    subtitle = "Cost per person per month over 12 months",
    x = "Months Since Diagnosis",
    y = "Cost PPPM ($)"
  ) +
  theme(panel.grid.minor = element_blank())
```

## Testing Edge Cases

When developing with Eunomia, test these scenarios:

```{r edge-cases}
# Test 1: Empty cohort handling
DatabaseConnector::executeSql(connection, "
  CREATE TABLE main.empty_cohort AS
  SELECT 
    1 AS cohort_definition_id,
    person_id AS subject_id,
    CURRENT_DATE AS cohort_start_date,
    CURRENT_DATE AS cohort_end_date
  FROM main.person
  WHERE 1 = 0  -- No results
")

emptyResult <- tryCatch({
  calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "empty_cohort",
    cohortId = 1,
    verbose = FALSE
  )
}, error = function(e) {
  cli::cli_alert_warning("Empty cohort handled gracefully: {e$message}")
  NULL
})

# Test 2: Very short time windows
shortWindowResult <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "gi_bleed_cohort",
  cohortId = 1,
  anchorCol = "cohort_start_date",
  startOffsetDays = 0,
  endOffsetDays = 1,  # Just 2 days
  verbose = FALSE
)

cli::cli_alert_info("Short window result - Patients with costs: {shortWindowResult$n_persons}")

# Test 3: Multiple event filters
complexFilters <- list(
  list(name = "Conditions", domain = "Condition", conceptIds = c(192671, 4112343)),
  list(name = "Procedures", domain = "Procedure", conceptIds = c(4019977, 4301351)),
  list(name = "Drugs", domain = "Drug", conceptIds = c(19122137, 1367500)),
  list(name = "Measurements", domain = "Measurement", conceptIds = c(3000963, 3004249))
)

complexResult <- calculateCostOfCare(
  connection = connection,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTable = "gi_bleed_cohort",
  cohortId = 1,
  eventFilters = complexFilters,
  verbose = TRUE
)
```

## Performance Tips for Eunomia

### 1. Use Indexes for Better Performance

```{r performance}
# Create useful indexes
DatabaseConnector::executeSql(connection, "
  CREATE INDEX IF NOT EXISTS idx_cost_person ON main.cost(person_id)
")

DatabaseConnector::executeSql(connection, "
  CREATE INDEX IF NOT EXISTS idx_cost_event ON main.cost(cost_event_id, cost_domain_id)
")
```

### 2. Profile Your Queries

```{r profiling}
# Example: Time different approaches
system.time({
  result1 <- calculateCostOfCare(
    connection = connection,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "gi_bleed_cohort",
    cohortId = 1,
    verbose = FALSE
  )
})
```

## Creating Reusable Test Scenarios

Build a library of test scenarios:

```{r test-scenarios}
# Function to create standard test cohorts
createStandardTestCohorts <- function(connection) {
  
  testCohorts <- list(
    smallCohort = "SELECT person_id FROM main.person LIMIT 100",
    mediumCohort = "SELECT person_id FROM main.person LIMIT 1000",
    largeCohort = "SELECT person_id FROM main.person",
    youngPatients = "SELECT person_id FROM main.person WHERE year_of_birth > 1980",
    oldPatients = "SELECT person_id FROM main.person WHERE year_of_birth < 1960",
    frequentVisitors = "
      SELECT person_id 
      FROM main.visit_occurrence 
      GROUP BY person_id 
      HAVING COUNT(*) > 10
    "
  )
  
  # Create each test cohort
  cohortId <- 1
  for(cohortName in names(testCohorts)) {
    sql <- sprintf("
      INSERT INTO main.test_cohorts
      SELECT 
        %d as cohort_definition_id,
        person_id as subject_id,
        DATE('2010-01-01') as cohort_start_date,
        DATE('2010-12-31') as cohort_end_date
      FROM (%s)
    ", cohortId, testCohorts[[cohortName]])
    
    DatabaseConnector::executeSql(connection, sql)
    cohortId <- cohortId + 1
  }
  
  cli::cli_alert_success("Created {length(testCohorts)} test cohorts")
}
```

## Clean Up

```{r cleanup}
# Remove temporary tables
tables_to_drop <- c(
  "condition_cohorts", "drug_cohorts", "gi_bleed_cohort", 
  "empty_cohort", "test_cohorts"
)

for(table in tables_to_drop) {
  DatabaseConnector::executeSql(
    connection, 
    sprintf("DROP TABLE IF EXISTS main.%s", table)
  )
}

# Disconnect
DatabaseConnector::disconnect(connection)
cli::cli_alert_success("Cleaned up and disconnected!")
```

## Summary

You've learned how to:

- ðŸ”§ Set up Eunomia with synthetic cost data
- ðŸ“Š Run various cost analyses on test data
- ðŸ§ª Test edge cases and performance
- ðŸ“ˆ Create visualizations of results
- ðŸ”„ Build reusable test scenarios

Eunomia is your sandbox - experiment freely and build confidence before working with real data!

## Next Steps

1. Try modifying the `injectCostData()` parameters to create different cost distributions
2. Experiment with complex event filters
3. Test performance with larger cohorts
4. Create your own test scenarios

Happy testing! ðŸš€